---
title: "Learn GEE"
author: "Joseph Holler"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lean GEE

-   Avoid treating observations from the same "cluster" as independent
-   Examples of "clusters" include:
    -   repeated measurements for the same person over time
    -   surveys of multiple members of the same household
    -   multiple observations in the same geographic region?
-   Some approaches to resolving this problem include
    -   discarding repeated observations
    -   including repeated observations but calculating statistical power based on the number of clusters rather than the number of observations
-   How can you use all the data points while also not exaggerating your statistical power?
-   GEE uses "weighted combinations of observations to extract the appropriate amount of information from correlated data"

Elements of notation:

|        Variables        | Parameter Symbol | Statistic Symbol |
|:-----------------------:|:----------------:|:----------------:|
|        household        |       $h$        |                  |
|          mean           |      $\mu$       |  $\overline{y}$  |
|   standard deviation    |     $\delta$     |                  |
|       proportion        |       $P$        |       $p$        |
| regression coefficient  |       $B$        |       $b$        |
| correlation coefficient |       $R$        |       $r$        |

## Variance of Weighted Sum

1.  Make a correlation matrix out of the variable and its weights
2.  For each row/column combination, find: row-weight \* column-weight \* row-δ \* column-δ \* row-column-$R$
3.  Sum the products from step 2

- If there is no correlation, then off-diagonal R values are 0, and the weighted variance is simply the sum of squared weights and variances.
- standard error of a mean is: $\mu$ / $\sqrt{n}$  ... and this is what you get if there is no correlation and equal weights in the sample
- standard error of an estimate is basically the standard deviation of it... 
- if data is correlated, the standard error increases (since you're no longer multiplying by 0 off-diagonal)

## What if some observations are clustered?

- three children: 1 single child and 2 siblings
- optimal weight for observations correlated within clusters is: 1 / (1 + $R$($k$-1))  where $k$ is the size of the cluster
- evidence that this is the optimal weight is shown in Figure 3 and described as provable with calculus on pg 368
- as correlation within clusters approaches 1, the effective sample size of the study decreases to the number of clusters. This make sense, as perfect correlation with clusters means that a cluster is effectively a single observation

## A challenge will be to estimate the global mean and global correlation

Here is a process for the estimation:

- start with assumption of no correlation $R$ = 0
- calculate estimate of $\mu$ based on $R$ = 0
- calculate new estimate $r$ of $R$
- repeat until you reach convergence (no more change in $R$)

```{r fig4}
library("tidyverse")
val <- c(15, 13, 10, 9, 8)
grp <- c(1, 1, 2, 2, 2)

# add code to count k members of each group

fig4data <- data.frame(val, grp)
fig4data <- fig4data %>% mutate(resid = val - mean(fig4data$val))
resid <- fig4data$val - mean(fig4data$val)

# estimated variance is sum of diagonal products divided by n(5) - 1
v <- sum(resid^2) / (length(resid) - 1)

# estimated covariance is sum of off-diagonal products divided by n(4) - 1
c <- (resid[1] * resid[2] + 
      resid[3] * resid[4] +
      resid[3] * resid[5] +
      resid[4] * resid[5] 
      ) / 3

# estimated correlation is estimated covariance / estimated variance
r <- c / v

# new weight is 1 / (1 + r(k-1))
fig4data <- fig4data %>% mutate(s2w <- 1 / (1 + r(fig4data$k - 1)))
```


